---
title: "Clustering"
output:
  word_document: default
  html_document: default
date: '2022-07-01'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```{r}
# Load necessary libraries
library(haven)
library(dendextend)
```


```{r}
# Load dataset from SPSS file
sizofrenii <- read_sav("wuschiz_1.sav")
```



```{r}
# Correlation Matrix
library("Hmisc")
rcorr(as.matrix(sizofrenii[c(3,4,6,8,9,11,13,15)])) 
# Table including the significance values (p-values) of the correlation matrix
```


```{r}
# Hierarchical Clustering
d <- dist(sizofrenii[c(3,4,6,8,9,11,13,15)], method = "euclidean") 
fit <- hclust(d, method="ward.D") 
dend <- as.dendrogram(fit)  # Create dendrogram
plot(dend)
plot(color_branches(dend, k=3))
```

```{r}
# Enhance dendrogram visual
geneldend <- dend %>%
  set("branches_lwd", 2) %>%
  set("branches_k_color", k = 3) %>%
  set("labels_cex", 1.2) %>%
  set("labels_colors", k=3)
plot(geneldend, font.axis = 2) 
```
```{r}
# Bootstrapped hierarchical clustering with pvclust
library(pvclust)
set.seed(13134)
result <- pvclust(sizofrenii[1:402, c(3,4,6,8,9,11,13,15)], 
                  method.dist = "cor", method.hclust = "average", nboot = 8)
plot(result)
pvrect(result)
# According to the dendrogram, the 'lie_scale' variable is further from other variables.
# Other variable clusters are closer to each other.
```
```{r}
# Visualize pvclust significance
dend %>% pvclust_show_signif(result) %>% 
   plot(main = "Cluster dendrogram \n bp values are highlighted by signif")

dend %>% pvclust_show_signif(result, show_type = "lwd") %>% 
   plot(main = "Cluster dendrogram with AU/BP values (%)\n bp values are highlighted by signif")

result %>% text
result %>% pvrect(alpha=0.95)

dend %>% pvclust_show_signif_gradient(result) %>% 
   plot(main = "Cluster dendrogram with AU/BP values (%)\n bp values are colored by signif")

dend %>%
   pvclust_show_signif_gradient(result) %>%
   pvclust_show_signif(result) %>%
   plot(main = "Cluster dendrogram with AU/BP values (%)\n bp values are colored+highlighted by signif")

result %>% text
result %>% pvrect(alpha=0.95)
```


```{r}
# Determining the optimal number of clusters (Elbow method)
library("factoextra")
fviz_nbclust(sizofrenii[c(3,4,6,8,9,11,13,15)], kmeans, method = "wss")
# Break point in the graph suggests 3 clusters is optimal.
```
```{r}
# Apply K-means clustering
set.seed(95739487) 
km.res <- kmeans(sizofrenii[c(3,4,6,8,9,11,13,15)], 3, iter.max = 20, algorithm = "Lloyd") 
t(km.res$centers)

# The table shows the mean values of each variable per cluster.
# Mixed high/low values across variables should be interpreted with caution.
```
```{r}
# Cluster visualization in 2D
library(cluster)
clusplot(sizofrenii[c(3,4,6,8,9,11,13,15)], km.res$cluster, 
         main = '2D representation of the Cluster solution',
         color = TRUE, shade = TRUE,
         labels = 2, lines = 0)
```
```{r}
# Alternative visualization using factoextra
fviz_cluster(km.res, data = sizofrenii[c(3,4,6,8,9,11,13,15)], palette = "jco",
             ggtheme = theme_minimal())
```
```{r}
# Add cluster labels to dataset
sizofrenii$cluster <- km.res$cluster
sizofrenii$cluster <- as.factor(sizofrenii$cluster)
```
```{r}
# ANOVA Tests to assess differences between clusters
library(rstatix)

yalan_olcegi_aov <- aov(yalan_olcegi ~ cluster, data = sizofrenii)
summary(yalan_olcegi_aov)

f_aov <- aov(f ~ cluster, data = sizofrenii)
summary(f_aov)

melankoli_aov <- aov(melankoli ~ cluster, data = sizofrenii)
summary(melankoli_aov)

histeri_aov <- aov(histeri ~ cluster, data = sizofrenii)
summary(histeri_aov)

psikopati_aov <- aov(psikopati ~ cluster, data = sizofrenii)
summary(psikopati_aov)

paranoya_aov <- aov(paranoya ~ cluster, data = sizofrenii)
summary(paranoya_aov)

Sc_aov <- aov(Sc ~ cluster, data = sizofrenii)
summary(Sc_aov)

ice_dönüklük_aov <- aov(ice_dönüklük ~ cluster, data = sizofrenii)
summary(ice_dönüklük_aov)

# Hypotheses:
# H0: There is no significant difference in variable means across clusters.
# H1: There is a significant difference in variable means across clusters.
# Since p < 0.05 for all variables, H0 is rejected → significant differences exist across clusters.
```
```{r}
# Filter each cluster to examine group sizes
sizofreni_1 <- sizofrenii %>% filter(cluster == 1)  # Cluster 1: 211 observations
sizofreni_2 <- sizofrenii %>% filter(cluster == 2)  # Cluster 2: 127 observations
sizofreni_3 <- sizofrenii %>% filter(cluster == 3)  # Cluster 3: 64 observations
```
